### 2016-01-18 papers/CAFe/HIPS-paper/coarray-extensions.tex
spelling

However, CAFe allows images to execute tasks on hosted subimages using standard procedure
calls.  For example,
%%\small
\begin{verbatim}
  call relax(U[device])  [[device]]
\end{verbatim}
%%\normalsize
executes the subroutine relax, on the subimage device, using array \texttt{U} \emph{located}
on the same subimage.  This follows the same pattern as introduced for remote coarray allocation,
whereby the double square bracket notation indicates \emph{where} the function will execute.
Funtions to be executed on a subimage must be pure procedures (i.e., declared with the Fortran
keyword \texttt{pure}) and scalar formal parameters must be passed by value (i.e., declared with the
\texttt{value} attribute).

>>>>

However, CAFe allows images to execute tasks on hosted subimages using standard procedure
calls.  For example,
%%\small
\begin{verbatim}
  call relax(U[device])  [[device]]
\end{verbatim}
%%\normalsize
executes the subroutine relax, on the subimage device, using array \texttt{U} \emph{located}
on the same subimage.  This follows the same pattern as introduced for remote coarray allocation,
whereby the double square bracket notation indicates \emph{where} the function will execute.
Functions to be executed on a subimage must be pure procedures (i.e., declared with the Fortran
keyword \texttt{pure}) and scalar formal parameters must be passed by value (i.e., declared with the
\texttt{value} attribute).

### 2016-01-18 papers/CAFe/HIPS-paper/conclusions.tex
spelling

Fortran is a general-purpose programming language and is extremely important to the
scientific and high performance computing (HPC) community.  While new code development is
often in languages other than Fortran (such as C++), surveys show that Fortran usage is
still high\cite{prabhu2011survey}.  Fortran will be with us well into the future as the
lifetime of important scientific applications is measured in decades and Fortran usage
dominates in scientific communitities that are critical to understanding the future of
life on our planet such as the climate community.

@@

\begin{itemize}
\item
  Dynamic creation of subimages.
\item
  Dynamic memory allocation and placement on a subimage.
\item
  Explicit addressing of memory on subimages using standard coarray notation.  It follows
  the Partitioned Global Address Space (PGAS) model, although with subimages the address
  space is hierahical in the sense that the memory partioning only allows memory exchange
  between a subimage and its hosting image as initiated by the hosting image.
\item
  Task creation and execution on subimages using extended coarray syntax with double
  square brackets \texttt{[[ ]]}.  Task mays be defined in terms of standard \emph{pure}
  Fortran procedures and task execution continues without interuption until completion.
  In addition, a Fortran 2015 event type may be notified upon completion of the task to allow
  concurrent execution on the subimage and on the hosting image.
\item
  CAFe integrates Fortran 2015 features to allow relatively simple programs to be written
  that employ all of the heterogenous components expected on exascale platforms.
\end{itemize}

@@

\begin{itemize}
\item
  The source-to-source transformations that have been developed to implement CAFe
  are rudimentory and have not been optimized.  For example, memory transfers have
  not been aggregated into larger message blocks to improve performance.  Memory transfer
  is also blocking so that prefetching of data cannot be employed via techniques
  envolving the reordering of statements (code motion) that are effectively employed in
  CAF compilers.
\item
  The Laplacian example is in reality a ``toy'' problem.  Real multiphysics codes
  employ hundreds (or more) of variables and would have entirely different performance
  characteristics because of the increased pressure on cache and register usage.
%\item
%  Existing scientific codes often separate communication into separate modules from
%  the computational kernels.  This makes adoption of CAF itself slow (because these
%  MPI modules already work fine) and thus additional extensions like CAFe also
%  would be slow to adopt. However CAFe constructs can be used in an incremental
%  fashion making it easier to adopt
%\item
%  In reality CAF (and therefore CAFe) are relatively low-level parallel programming
%  languages that require the programmer to \emph{explicitly} control the placement
%  of memory, communication, and synchronization.  
\item
  Thorough examination of performance capabilities on real scientific applications
  (or at least scientific ``mini-apps'') has not been attempted.  Such a study of both
  CAFe and CAF itself is necessary to demonstrate the coarray model as a viable alternative
  for application codes with complexity above that of an idealized mini-app.
\item
  Programming models like that proposed by CAFe already exist in the form of a combination
  of MPI and OpenMP (or OpenACC) programs.  The advantage of CAFe --- by being a purely
  language construct --- is that it opens up the possibility of compiler optimizations
  that would otherwise be impossible with a library approach intermixed with compiler directives.
\end{itemize}

@@

In spite of the limitations discussed, 
adoption of the task-based parallelism of CAFe and the synchronization
event types in Fortran 2015 suggests intriguing possibilities.  Tasks plus the
dependencies represented by events suggest a correspondence with the Open Community
Runtime (OCR) libraries.  In OCR, the entire program must be broken into a set of tasks
represented by a directed acycic graph (DAG).  In OCR, tasks are available for executing
when all of their input dependencies are satisfied, such as occurs when computation of a new
iteration of data elements have been updated and other upstream tasks have been completed.

Tools to automatically convert existing applications to run under the OCR model faces challenges.  For
example, the hundreds of MPI functions would have to be semantically understood by the tools,
integrated with existing OpenACC directives (for example), a DAG generaged for the complete
program, and then transformed into an OCR program.

>>>>

Fortran is a general-purpose programming language and is extremely important to the
scientific and high performance computing (HPC) community.  While new code development is
often in languages other than Fortran (such as C++), surveys show that Fortran usage is
still high\cite{prabhu2011survey}.  Fortran will be with us well into the future as the
lifetime of important scientific applications is measured in decades and Fortran usage
dominates in scientific communities that are critical to understanding the future of
life on our planet such as the climate community.

@@

\begin{itemize}
\item
  Dynamic creation of subimages.
\item
  Dynamic memory allocation and placement on a subimage.
\item
  Explicit addressing of memory on subimages using standard coarray notation.  It follows
  the Partitioned Global Address Space (PGAS) model, although with subimages the address
  space is hierarchical in the sense that the memory partitioning only allows memory exchange
  between a subimage and its hosting image as initiated by the hosting image.
\item
  Task creation and execution on subimages using extended coarray syntax with double
  square brackets \texttt{[[ ]]}.  Task mays be defined in terms of standard \emph{pure}
  Fortran procedures and task execution continues without interruption until completion.
  In addition, a Fortran 2015 event type may be notified upon completion of the task to allow
  concurrent execution on the subimage and on the hosting image.
\item
  CAFe integrates Fortran 2015 features to allow relatively simple programs to be written
  that employ all of the heterogenous components expected on exascale platforms.
\end{itemize}

@@

\begin{itemize}
\item
  The source-to-source transformations that have been developed to implement CAFe
  are rudimentary and have not been optimized.  For example, memory transfers have
  not been aggregated into larger message blocks to improve performance.  Memory transfer
  is also blocking so that prefetching of data cannot be employed via techniques
  involving the reordering of statements (code motion) that are effectively employed in
  CAF compilers.
\item
  The Laplacian example is in reality a ``toy'' problem.  Real multiphysics codes
  employ hundreds (or more) of variables and would have entirely different performance
  characteristics because of the increased pressure on cache and register usage.
%\item
%  Existing scientific codes often separate communication into separate modules from
%  the computational kernels.  This makes adoption of CAF itself slow (because these
%  MPI modules already work fine) and thus additional extensions like CAFe also
%  would be slow to adopt. However CAFe constructs can be used in an incremental
%  fashion making it easier to adopt
%\item
%  In reality CAF (and therefore CAFe) are relatively low-level parallel programming
%  languages that require the programmer to \emph{explicitly} control the placement
%  of memory, communication, and synchronization.  
\item
  Thorough examination of performance capabilities on real scientific applications
  (or at least scientific ``mini-apps'') has not been attempted.  Such a study of both
  CAFe and CAF itself is necessary to demonstrate the coarray model as a viable alternative
  for application codes with complexity above that of an idealized mini-app.
\item
  Programming models like that proposed by CAFe already exist in the form of a combination
  of MPI and OpenMP (or OpenACC) programs.  The advantage of CAFe --- by being a purely
  language construct --- is that it opens up the possibility of compiler optimizations
  that would otherwise be impossible with a library approach intermixed with compiler directives.
\end{itemize}

@@

In spite of the limitations discussed, 
adoption of the task-based parallelism of CAFe and the synchronization
event types in Fortran 2015 suggests intriguing possibilities.  Tasks plus the
dependencies represented by events suggest a correspondence with the Open Community
Runtime (OCR) libraries.  In OCR, the entire program must be broken into a set of tasks
represented by a directed acyclic graph (DAG).  In OCR, tasks are available for executing
when all of their input dependencies are satisfied, such as occurs when computation of a new
iteration of data elements have been updated and other upstream tasks have been completed.

Tools to automatically convert existing applications to run under the OCR model faces challenges.  For
example, the hundreds of MPI functions would have to be semantically understood by the tools,
integrated with existing OpenACC directives (for example), a DAG generated for the complete
program, and then transformed into an OCR program.

### 2016-01-18 papers/CAFe/HIPS-paper/multigrid.tex
spelling

In this section a one-dimensional solution to Laplace's equation is described using CAFe
syntax.  The solution uses multigrid techniques to improve the rate of convergence
of iterative methods.  We decribe the implementation in terms of a one-dimensional
problem for simplicity.

@@

The multigrid algorithm uses a series of successively coarser grids to
iteratively approximate the solution obtained from the preceeding
finer grid.  Higher frequency error modes are damped out as the
solution processes up the grid hierarchy.  At the top of the
hierarchy, an exact solution is obtained and then this solution is
iterated and propogated down the grid hierarchy form coarser to finer
grids.  After a few sweeps up and down the hierarchy the solution with
have been obtained on the desired original grid.

@@

Two levels of the grid hierarchy are shown in \ref{fig_grids} running on a processor computing on
an interior portion of the grid; additional images are computing on regions to the left and right of
the region shown in the figure.  The finist grid (size N) is shown
in the bottom of the figure with cell indices running from (-1:N+1).  Cells 0 and N are shared
between image neighbors to the left and right (indicated by the light blue coloring and the vertical
dashed lines).  These shared cells are computed redundantly on each program image.

@@

At the top of the grid hierarchy there are sufficiently few cells that an exact solution can be
obtained (perhaps by a non-iterative method).  This solution is then propagated down the
grid hierarchy as shown in Figure\ref{fig_prolongate}.

@@

The multigrid algorithm has been implemented in Fortran using the CAFe
syntax described in the previous section. Data coarrays at each grid
level are declared, allocated, and initialized on each program image.
These coarrays are then copied to the subimages by the parent, hosting
images.  Below we show the implementation of the relaxation loop on
grid level N.  Each iteration step incorporates concurrent computation
on an image and its hosted subimage.  In order to execute
concurrently, we must declare a event variable to allow concurrent
execution and then synchronization once \emph{both} computation ond
the subimage \emph{and} the image have completed.

The event variable is declared as,
\small
\begin{verbatim}
 type(event_type) :: evt
\end{verbatim}
\normalsize
After initialization, this event variable has its count variable set to 0 and is incremented
each time the subimage task completes during an interation of the code segment shown below:
\small
\begin{verbatim}
 do i = 1, nsteps                            1
   call relax (N, V1h[dev], Buf[dev])        &
                    [[dev, EVENT=evt]]       2
   call relax_boundary (N, V1h)              3

@@

Most of the computation is accomplised by the subimage executing the \texttt{relax}
procedure (not shown) operating over the interior corresponding to fine grid indices (1:N-1).
Concurrently, \texttt{relax\_boundary} operates on the boundary indices 0 and N
(note that memory for the boundary computation in located on local image as
the coarray \texttt{V1h} is not coindexed using square bracket notation in statement 3.

>>>>

In this section a one-dimensional solution to Laplace's equation is described using CAFe
syntax.  The solution uses multigrid techniques to improve the rate of convergence
of iterative methods.  We describe the implementation in terms of a one-dimensional
problem for simplicity.

@@

The multigrid algorithm uses a series of successively coarser grids to
iteratively approximate the solution obtained from the preceding
finer grid.  Higher frequency error modes are damped out as the
solution processes up the grid hierarchy.  At the top of the
hierarchy, an exact solution is obtained and then this solution is
iterated and propagated down the grid hierarchy form coarser to finer
grids.  After a few sweeps up and down the hierarchy the solution with
have been obtained on the desired original grid.

@@

Two levels of the grid hierarchy are shown in \ref{fig_grids} running on a processor computing on
an interior portion of the grid; additional images are computing on regions to the left and right of
the region shown in the figure.  The finest grid (size N) is shown
in the bottom of the figure with cell indices running from (-1:N+1).  Cells 0 and N are shared
between image neighbors to the left and right (indicated by the light blue coloring and the vertical
dashed lines).  These shared cells are computed redundantly on each program image.

@@

At the top of the grid hierarchy there are sufficiently few cells that an exact solution can be
obtained (perhaps by a non-iterative method).  This solution is then propagated down the
grid hierarchy as shown in Figure \ref{fig_prolongate}.

@@

The multigrid algorithm has been implemented in Fortran using the CAFe
syntax described in the previous section. Data coarrays at each grid
level are declared, allocated, and initialized on each program image.
These coarrays are then copied to the subimages by the parent, hosting
images.  Below we show the implementation of the relaxation loop on
grid level N.  Each iteration step incorporates concurrent computation
on an image and its hosted subimage.  In order to execute
concurrently, we must declare a event variable to allow concurrent
execution and then synchronization once \emph{both} computation and
the subimage \emph{and} the image have completed.

The event variable is declared as,
\small
\begin{verbatim}
 type(event_type) :: evt
\end{verbatim}
\normalsize
After initialization, this event variable has its count variable set to 0 and is incremented
each time the subimage task completes during an iteration of the code segment shown below:
\small
\begin{verbatim}
 do i = 1, nsteps                            1
   call relax (N, V1h[dev], Buf[dev])        &
                    [[dev, EVENT=evt]]       2
   call relax_boundary (N, V1h)              3

@@

Most of the computation is accomplished by the subimage executing the \texttt{relax}
procedure (not shown) operating over the interior corresponding to fine grid indices (1:N-1).
Concurrently, \texttt{relax\_boundary} operates on the boundary indices 0 and N
(note that memory for the boundary computation in located on local image as
the coarray \texttt{V1h} is not coindexed using square bracket notation in statement 3.
